---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---


## Loading and preprocessing the data

The code below sets some variables concerning the script enviroment in order
to standardize and facilitate its use.  
It also loads some general libraries that will be used on the following steps.

```{r echo = TRUE}
library(dplyr)
library(plyr)

workingdirectory <- "~/R/RepData_PeerAssessment1"
zipfile.name <- "activity.zip"
csvfile.name <- "activity.csv"
files.path <- workingdirectory
zipfile.path <- paste(files.path, zipfile.name, sep = "/")
csvfile.path <- paste(files.path, csvfile.name, sep = "/")

setwd(workingdirectory)
```

As the next step it unzips and loads the dataset. In order to do that, it 
verifies if the file is where expected (the zip file dataset must be provided 
and placed into the specific folder as defined into the **zipfile.path** variable).
The script also checks if load/unzip step is really required once it may have 
been executed beforehands.
```{r echo = TRUE}
# Check if zipfile is where expected
if (!file.exists(zipfile.path)) {
    stop(
        paste(
            "Message: The script couldn't continous because the data file was not found at '",
            zipfile.path, "'", ". Solution: place the ", zipfile.name,
            " (the one that contains the data source for this script) at ",
            files.path, " directory and try again", sep = ""
        )
    )
}

# Unzip if csv file doesn't exists
if (!file.exists(csvfile.path)) {
    unzip(zipfile.path)
}

# Read the data
activity.rawdata <-
    read.csv(
        file = csvfile.path, header = TRUE, sep = ",", stringsAsFactors = FALSE
    )
# Create a different dataframe so the raw data is preserved at its original 
# format in order to future need
activity.data <- activity.rawdata

```

From now on, it formats the columns to the correct type of value.

```{r echo = TRUE}

# Converts steps columns into number values
if (class(activity.data$steps) != "integer") {
    activity.data$steps <- as.numeric(activity.data$steps)
}

# Converts the date column into date values
if (class(activity.data$date) != "Date") {
    activity.data$date <-
        as.Date(x = as.character(activity.data$date), format = "%Y-%m-%d")
}

# Converts the interval column into number values
if (class(activity.data$steps) != "integer") {
    activity.data$steps <- as.numeric(activity.data$steps)
}
```

The code below just brings up some additional information about the dataframe
for you to be familiar with its content.

```{r echo = TRUE}
# Data structure info
str(activity.data)

# Percental of NA at each column
sapply(
    activity.data, FUN = function(x) {
        mean(is.na(x))
    }
)
```

## What is the mean total number of steps taken per day?

To answer this question, the script executes some preparation steps in order
to preserve the cleanned and tidied dataset that was built previously.
Thus some new variables were created and some additional cleanning data steps
executed, since all NA values will be ignored to answer the currently.

```{r echo = TRUE}
# Create a new dataset with no NA value
activity.naignored <- activity.data[!is.na(activity.data$steps),]
```

The following code summarises the total number of steps for each day and 
plots the result into a histogram. Furthermore, it scratches a vertical blue line
into the graph indicating the average number of total taken steps per day.

```{r echo = TRUE}
# Summarises the total daily number os steps
totalbyday.naignored <- ddply(
    activity.naignored, 
    .(date),
    summarise,
    total = sum(steps)
)

# Plots the histogram
hist(totalbyday.naignored$total,
     main = "Histogram of the total number of steps taken each day", 
     xlab = "Number of the steps in one day",
     ylab = "Frequency of days"
)

# Adds a vertical blue line indicating the daily average number of steps
abline(v = mean(totalbyday.naignored$total), lwd = 3, col = "blue")

```

Besides the graph information below, the following code reports some numerical 
info concerning what has been asked.

```{r echo = TRUE}
# mean number of steps taken each day
mean(totalbyday.naignored$total)

# median number of steps taken each day
median(totalbyday.naignored$total)
```

As you can see, the average number of total daily steps is
`r mean(totalbyday.naignored$total)`.

## What is the average daily activity pattern?

To answer this question the code below took as its input the data frame
with NA values ignored.
The first step was calculating the mean number of steps of each 5-minute interval
across all days.

```{r echo = TRUE}
# Calculate the mean number of steps of each 5-minute interval across all days
meanbyinterval.naignored <- ddply(
    activity.naignored, 
    .(interval), 
    summarise, 
    mean = mean(steps)
)
```

Then, it plotted the result into line graph.

```{r echo = TRUE}
#Plot the graph with the average number of steps for each 5-minutes interval
plot(
    x = meanbyinterval.naignored$interval, 
    y = meanbyinterval.naignored$mean, 
    type = "l", 
    xlab = "Time (5-minutes interval)", 
    ylab = "Number of steps", 
    main = "Average number of steps taken"
)
```

As a complementary info, the script identifies in which 5-minutes interval
the higgest average of total number of steps happen.

```{r echo = TRUE}
# Gets the 5-minutes interval which has the maximum average number of steps
# across all days
meanbyinterval.naignored[
    meanbyinterval.naignored$mean == max(meanbyinterval.naignored$mean),]
```

## Imputing missing values

As you could see in the beginning of this analises, the original dataset contains
NA values. For the previous analyses, theses observations were ignored, but 
it would be interesting to check if by ignoring all these values (
```r sum(!complete.cases(activity.data))``` to be more precise, or 
```r mean(!complete.cases(activity.data))*100```% of the total observations) we are not 
changing the result significantly. For this purpose the code created
a new data frame where no observation containing NA values were ignored and its
values replaced by its correspondent average number of total steps 
interval across all days.  
First, the code bellow reports the number and percental of observations 
containing NA values.

```{r echo = TRUE}
# Calculate the number and percental of rows with missing values
sum(!complete.cases(activity.data))

mean(!complete.cases(activity.data))

```

Now it iterates over the data frame and replaces all NA values by it's 
correspondent mean 5-minutes interval.

```{r echo = TRUE}
# Replace NA values by interval mean
activity.nafilled <- activity.data
for (i in 1:nrow(activity.nafilled)) {
        #Get the step value
        step.value <- activity.nafilled[i,"steps"]
        if (is.na(step.value)) {
            #Get the step interval
            step.interval <- activity.nafilled[i, "interval"]
            
            #Repacle step value from na by the mean step value at the correspond
            #interval
            activity.nafilled[i,"steps"] <- 
                meanbyinterval.naignored[
                    meanbyinterval.naignored$interval == step.interval, "mean"]
        }
}
# remove the variables that won't be used on the following code.
rm(step.value, step.interval)
```

Now it's time to report the results. First, the total daily number of steps
is ilustrated into a histogram which also contains one vertical blue line 
indicating the average number of daily total steps.

```{r echo = TRUE}
# Calculate the total daily number of steps from activity.nafilled variable
totalbyday.nafilled <- ddply(
    activity.nafilled, 
    .(date),
    summarise,
    total = sum(steps)
)

# Plots the histogram
hist(totalbyday.nafilled$total,
     main = "Frequency of days by its number of steps", 
     xlab = "Number of the steps in one day",
     ylab = "Frequency of days"
)

# Adds a vertical blue line that marks the daily average number of steps
abline(v = mean(totalbyday.nafilled$total), lwd = 3, col = "blue")
```

Now the numeric mean and median values of total number of steps taken per day.

```{r echo = TRUE}
# mean daily number of steps
mean(totalbyday.nafilled$total)

# median daily number of steps
median(totalbyday.nafilled$total)
```

Lastly the boxplot and the summary comparison below allow you to identify that
by ignoring or replacing NA values by its correspondent 5-minutes interval mean
doesn't make any big difference, once the mean value in both cases remains almost the
same. Some difference can be seen only at the 1st and 3rd quartiles.

```{r echo = TRUE}
# Plots a boxplot graph so it enables a visually comparison between ignored and 
# filled NAs values
totalbyday.nafilled <- cbind(totalbyday.nafilled, datasource = "na.filled")
totalbyday.naignored <- cbind(totalbyday.naignored, datasource = "na.ignored")
totalbyday.union <- rbind(totalbyday.nafilled, totalbyday.naignored) %>% 
    data.frame()
boxplot(
    total ~ datasource, 
    totalbyday.union, 
    main = "Total daily steps from Filled Vs Ignored NAs")

cbind(
    na.filled = summary(totalbyday.nafilled$total), 
    na.ignored = summary(totalbyday.naignored$total)
)
```

## Are there differences in activity patterns between weekdays and weekends?

To try to answer this question the code below identifies which is the weekday
of each sample. Then, it classified the days of the week into **weekend** or 
**weekday**.  
Next, it finds the mean number of steps by each 5-minutes interval across all 
days to finally plot all these data into one line graph.

```{r echo = TRUE}
# Add the weekday (sun-sat) and weekdaytype (weekday or weekend) columns
activity.weekday <- cbind(
    activity.nafilled, 
    weekday = sapply(activity.nafilled$date, weekdays)
    )
activity.weekday <- cbind(
    activity.weekday, 
    weekdaytype = sapply(activity.weekday$weekday, function(x){
        if(as.numeric(x) %in% c(2,3,5,6,7) ){
            "weekday"
        }else{
            "weekend"
        }
    })
    ) %>% 
    data.frame()

# Calculate the mean number os steps by each 5-minutes interval
total.byintervalandweekday <- ddply(
    activity.weekday, 
    .(interval, weekdaytype),
    summarise,
    mean = mean(steps)
)

# Plot hte values by
library(ggplot2)
ggplot(
    data = total.byintervalandweekday, 
    aes(x = interval, y = mean, color = factor(weekdaytype)),
    xlab = "Interval",
    ylab = "Mean number of steps"
) + geom_line()  + scale_color_hue(name = "Legend") + ggtitle(label = "Mean number of steps by interval - Weekend Vs Weekday")
```

By checking out the last graph it's possible to conclude:  
- People usually start walking after the 500th 5-minutes interval;  
- On the weekdays the number of steps at the initial intervals is higher thant 
those at the weekend;  
- For both, weekend and weekday, the number of steps between the 800th and 
1.000th (aprox.) interval, the number of steps seams to be the highest of the 
day, indicating a possible running activity during this period of time;  
- Along all the rest of the day people walk more during the weekends than during
the weekdays, problably because they are usually at the office, where it's more common
to be sitted than to be walking.